---
title: "Periodontal Microbiome Signatures"
format: 
    html:
        grid:
            sidebar-width: 50px
            body-width: 1000px
            margin-width: 300px
        toc: true
        toc-depth: 2 
        code-fold: true
        code-tools: true
        theme: 
          light: [flatly, style.scss]
          dark: [darkly, style.scss]
        fontsize: 1.2em
        linestretch: 2
        highlight-style: a11y
        date: "February 10, 2026"
        embed-resources: true
title-block-banner: "images/HMS_DBMI_Logo.png"
title-block-banner-color: "#333333"
css: style.css
bibliography: refs.bib
author:
    - name: Alex Pickering
      orcid: 0000-0002-0002-6759
      email: alex_pickering@hms.harvard.edu
      affiliation:
          - name: CCB
            url: https://ccb.hms.harvard.edu/
            id: ccb
      roles: [formal analysis, software]
    - name: Tram Nguyen
      orcid: 0000-0003-4809-6227
      email: tram_nguyen@hms.harvard.edu
      affiliation:
          - name: CCB
            url: https://ccb.hms.harvard.edu/
            id: ccb
      roles: [formal analysis, software]
    - name: Ludwig Geistlinger 
      orcid: 0000-0002-2495-5464
      email: ludwig_geistlinger@hms.harvard.edu
      affiliation:
          - ref: ccb
      roles: supervision
license: 
    - text: "HMS authorship guidelines: https://hms.harvard.edu/sites/default/files/assets/Sites/Ombuds/files/AUTHORSHIP%20GUIDELINES.pdf"
    - url: https://hms.harvard.edu/sites/default/files/assets/Sites/Ombuds/files/AUTHORSHIP%20GUIDELINES.pdf
editor: source 
---
 
This report was rendered on `r format(Sys.time(), "%B %d, %Y")`.



```{r}
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(bugSigSimple)
library(ggplot2)
library(ComplexHeatmap)
library(circlize)
library(dendextend)

# TODO: replace by bigsigdbr::importBugSigDB()
dat <- readRDS('example_data/perio_bugs.rds')
```

# Background
Periodontitis is a chronic inflammatory disease affecting the supporting structures of teeth, including the gingiva, periodontal ligament, and alveolar bone. The disease is driven by dysbiosis of the oral microbiome, with shifts from commensal to pathogenic bacterial communities. Understanding which microbial taxa are consistently associated with periodontitis across multiple studies can help identify key drivers of disease and potential therapeutic targets.

PerioSigDB is a comprehensive database of periodontal microbial signatures curated from published literature, providing a resource for meta-analysis of microbiome studies. Here, we analyze periodontal disease-associated signatures from PerioSigDB to identify taxa that show consistent patterns of abundance changes across studies.

# Taxa Increased in Periodontitis

From the periodontal microbiome signatures, we generated a table of taxa showing consistent increases in periodontitis. The binomial test evaluates whether each taxon is more frequently reported as increased rather than decreased across studies. Taxa appearing in this table show statistically significant directional consistency.

```{r}
#| classes: plain

dat_condition <- dat |> 
  mutate(comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs "))

# table of studies
taxon_table <- bugSigSimple::createTaxonTable(dat_condition, n=20) |> 
  select(
    `Taxon Name`,
    `Signatures` = total_signatures,
    `Increased` = increased_signatures,
    `Decreased` = decreased_signatures,
    `P-value` = `Binomial Test pval`
  )

taxon_table |> 
  dplyr::filter(Increased > Decreased) |>
  dplyr::filter(`P-value` < 0.05) |>
  kableExtra::kbl()
```

# Taxa Decreased in Periodontitis

Conversely, we identified taxa that are consistently depleted in periodontitis compared to healthy controls. These may represent commensal or health-associated microbes that are displaced during disease progression.

```{r}
#| classes: plain

taxon_table |> 
  dplyr::filter(Decreased > Increased) |>
  kableExtra::kbl()

```

# Signature Length

Understanding the size distribution of microbial signatures provides insight into the complexity of microbiome alterations in periodontitis. Most signatures report relatively few differentially abundant taxa:

```{r}
#| fig-width: 10

# cluster analysis
allsigs <- bugsigdbr::getSignatures(dat_condition, tax.id.type = "taxname")
allsigs <- allsigs[sapply(allsigs, length) > 1] #require length > 1

# what is the distribution of signature lengths?
siglengths <- sapply(allsigs, length)
siglengths.df <- data.frame(siglengths = siglengths)
ggplot(siglengths.df, aes(x=siglengths)) +
  geom_bar(width=.9, colour='black', fill= '#ddd', linewidth=0.5) +
  theme_bw() +
  xlab('Number of Microbes in Signature') +
  ylab('Number of Signatures')

```

The distribution shows that most signatures contain between 1-10 taxa, with a long tail of larger signatures. This pattern may reflect differences in study design, sequencing depth, statistical power, or biological reality.

# Cluster Analysis

To assess consistency across studies, we calculated pairwise Jaccard similarities between all signatures. The heatmap below shows how similar signatures are to each other, with hierarchical clustering grouping similar signatures together. The top annotation bar shows the size (number of taxa) of each signature.

```{r}
#| fig-width: 10
#| fig-height: 8
mydists <- BugSigDBStats::calcPairwiseOverlaps(allsigs)

# Create a matrix of Jaccard similarities (0 for no overlap, 1 for 100% overlap)
jmat <- BugSigDBStats::calcJaccardSimilarity(allsigs)

# Remove annotation name, use column title on heatmap
ha <- HeatmapAnnotation(
  `Signature Length` = anno_barplot(siglengths),
  show_annotation_name = FALSE  # Hide the annotation label
)
hm <- Heatmap(
  jmat,
  top_annotation = ha, 
  column_title = "Signature Length",  # Add as column title
  column_title_side = "top",
  left_annotation = NULL,
  row_names_max_width = unit(20, "cm"),
  column_names_max_height = unit(20, "cm"),
  row_labels = gsub('^.+?_(UP|DOWN)$', "\\1", rownames(jmat)),
  column_labels = gsub('^.+?_(UP|DOWN)$', "\\1", colnames(jmat)),
  row_names_gp = gpar(fontsize = 6),  
  column_names_gp = gpar(fontsize = 6),
  show_column_dend = FALSE,
  heatmap_legend_param = list(
    title = 'Jaccard Similarity',
    direction = "horizontal"  # Make legend horizontal
  )
)

# Draw with legend at bottom
draw(hm, heatmap_legend_side = "bottom")
```

 Signatures reporting increased taxa (UP) in periodontitis show particularly strong clustering, suggesting a core set of disease-associated microbes is consistently identified.
 
# Explore Species Within Clusters
```{r}
#| fig-width: 10
#| fig-height: 8

# Helper function to get author labels (without year)
get_author_label <- function(dat, sname) {
  id <- unlist(strsplit(sname, "_"))[1]
  id <- gsub("^bsdb:", "", id)
  id <- unlist(strsplit(id, "/"))
  
  sdat <- dat[dat$Study == paste0("Study ", id[1]) & 
                dat$Experiment == paste0("Experiment ", id[2]), ]
  
  if(nrow(sdat) == 0) return("Unknown")
  
  author <- sdat$`Authors list`[1]
  # Get first author last name
  aspl <- unlist(strsplit(author, " "))[1]
  year <- sdat$Year[1]
  
  # Add "et al." if multiple authors
  if(grepl(",", author)) {
    return(paste0(aspl, " et al. (", year, ")"))
  } else {
    return(paste0(aspl, " (", year, ")"))
  }
}
# Helper function to plot frequency heatmaps
create_taxa_heatmap <- function(dat_condition, 
                                clust_up, 
                                jmat, 
                                hc, 
                                tax_level = "species") {
  
  # Get signatures
  sigs_taxa <- bugsigdbr::getSignatures(
    dat_condition,
    tax.id.type = "taxname",
    tax.level = tax_level,
    exact.tax.level = FALSE
  )
  
  # Filter to only signatures in clust_up
  sigs_taxa <- sigs_taxa[names(sigs_taxa) %in% clust_up]
  
  # Remove empty signatures
  sigs_taxa <- sigs_taxa[sapply(sigs_taxa, length) > 0]
  
  # Create binary presence/absence matrix
  all_taxa <- unique(unlist(sigs_taxa))
  
  taxa_matrix <- matrix(0, 
                        nrow = length(sigs_taxa),
                        ncol = length(all_taxa),
                        dimnames = list(names(sigs_taxa), all_taxa))
  
  for(i in seq_along(sigs_taxa)) {
    sig_name <- names(sigs_taxa)[i]
    taxa_in_sig <- sigs_taxa[[i]]
    taxa_matrix[sig_name, taxa_in_sig] <- 1
  }
  
  # Calculate taxa frequency
  taxa_freq <- colSums(taxa_matrix) / nrow(taxa_matrix)
  taxa_order <- order(taxa_freq, decreasing = TRUE)
  taxa_freq_ordered <- taxa_freq[taxa_order]
  
  # Re-cluster using original distance matrix
  sigs_to_use <- rownames(taxa_matrix)
  jmat_subset <- jmat[sigs_to_use, sigs_to_use]
  hc_subset <- hclust(dist(jmat_subset), method = hc$method)
  dend_subset <- as.dendrogram(hc_subset)
  
  # Reorder taxa_matrix to match clustering
  taxa_matrix_ordered <- taxa_matrix[hc_subset$labels, taxa_order]
  
  # Apply author labels
  sig_labels_ordered <- vapply(rownames(taxa_matrix_ordered), 
                               get_author_label, 
                               character(1), 
                               dat = dat_condition)
  
  labels(dend_subset) <- sig_labels_ordered
  rownames(taxa_matrix_ordered) <- sig_labels_ordered
  
  # Create color matrix
  color_matrix <- taxa_matrix_ordered
  for(j in 1:ncol(color_matrix)) {
    for(i in 1:nrow(color_matrix)) {
      if(color_matrix[i, j] == 0) {
        color_matrix[i, j] <- NA
      } else {
        color_matrix[i, j] <- taxa_freq_ordered[j]
      }
    }
  }
  
  # Color function
  col_fun <- colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
  
  # Create heatmap
  hm_taxa <- Heatmap(
    color_matrix,
    name = "Frequency",
    col = col_fun,
    na_col = "white",
    cluster_rows = dend_subset,
    cluster_columns = FALSE,
    show_row_dend = TRUE,
    row_dend_width = unit(1, "cm"),
    row_names_side = "left",
    column_names_side = "bottom",
    row_names_gp = gpar(fontsize = 10),
    column_names_gp = gpar(fontsize = 10),
    column_names_rot = 90,
    column_title = NULL,
    row_title = NULL,
    top_annotation = NULL,
    heatmap_legend_param = list(
      title = "Frequency",
      direction = "vertical",
      title_gp = gpar(fontsize = 9),
      labels_gp = gpar(fontsize = 8),
      at = c(0, 0.25, 0.5, 0.75, 1),
      labels = c("0", "0.25", "0.5", "0.75", "1"),
      legend_height = unit(2, "cm"),
      legend_width = unit(0.3, "cm")
    )
  )
  
  return(hm_taxa)
}
```

Another way to explore the species that are more abundant in periodontitis is to look at the species present within the studies whose signatures are highly similar (the cluster we identified in the previous section). We start but looking at signatures at the genus level:


```{r}
#| fig-width: 10
#| fig-height: 8
# identify the clusters
hc <- hclust(dist(jmat))

#set the number of clusters here with k
clusts <- sort(cutree(hc, k = 2)) 
clusts <- lapply(unique(clusts), function(i) names(clusts)[clusts == i])

# proceed with cluster of UP signatures with high similarity
clust_up <- clusts[[2]]

hm_genus <- create_taxa_heatmap(dat_condition, clust_up, jmat, hc, 
                                  tax_level = "genus")
draw(hm_genus, heatmap_legend_side = "right")
```

If we want a bit more detail we can also look at the same plot for the species level:

```{r}
#| fig-width: 10
#| fig-height: 8

hm_species <- create_taxa_heatmap(dat_condition, clust_up, jmat, hc, 
                                  tax_level = "species")
draw(hm_species, heatmap_legend_side = "right")

```
# References

:::{#refs}
:::