---
title: "Periodontal Microbiome Signatures"
format: 
    html:
        grid:
            sidebar-width: 50px
            body-width: 1100px
            margin-width: 300px
        toc: true
        toc-depth: 2 
        code-fold: true
        code-tools: true
        theme: 
          light: [flatly, style.scss]
          dark: [darkly, style.scss]
        fontsize: 1.2em
        linestretch: 2
        highlight-style: a11y
        date: "February 10, 2026"
        embed-resources: true
title-block-banner: "images/HMS_DBMI_Logo.png"
title-block-banner-color: "#333333"
css: style.css
bibliography: refs.bib
author:
    - name: Alex Pickering
      orcid: 0000-0002-0002-6759
      email: alex_pickering@hms.harvard.edu
      affiliation:
          - name: CCB
            url: https://ccb.hms.harvard.edu/
            id: ccb
      roles: [formal analysis, software]
    - name: Tram Nguyen
      orcid: 0000-0003-4809-6227
      email: tram_nguyen@hms.harvard.edu
      affiliation:
          - name: CCB
            url: https://ccb.hms.harvard.edu/
            id: ccb
      roles: [formal analysis, software]
    - name: Ludwig Geistlinger 
      orcid: 0000-0002-2495-5464
      email: ludwig_geistlinger@hms.harvard.edu
      affiliation:
          - ref: ccb
      roles: supervision
license: 
    - text: "HMS authorship guidelines: https://hms.harvard.edu/sites/default/files/assets/Sites/Ombuds/files/AUTHORSHIP%20GUIDELINES.pdf"
    - url: https://hms.harvard.edu/sites/default/files/assets/Sites/Ombuds/files/AUTHORSHIP%20GUIDELINES.pdf
editor: source 
---
 
This report was rendered on `r format(Sys.time(), "%B %d, %Y")`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r helper-functions}
library(tidyverse)
library(bugSigSimple)
library(ggplot2)
library(ComplexHeatmap)
library(circlize)
library(dendextend)

# Helper functions ----

get_most_specific <- function(taxpart) {
  taxpart_split <- strsplit(taxpart, split='|', fixed=TRUE)
  most_specific <- sapply(taxpart_split, tail, 1)
  sapply(strsplit(most_specific, '__'), tail, 1)
}

make_freq_dat <- function(dat_exp, direction, min_freq = 3) {
  dat_exp |>
    filter(`Abundance in Group 1` == direction) |> 
    count(last_taxname, last_taxid, name = "freq") |>
    filter(freq >= min_freq) |> 
    arrange(desc(freq)) |> 
    mutate(across(c(last_taxid, last_taxname), ~ factor(., levels = .)))
}

make_freq_dat_combined <- function(dat_exp, pval_dat, min_freq = 3) {
  dat_up_unfiltered <- dat_exp |>
    filter(`Abundance in Group 1` == "increased") |> 
    count(last_taxname, last_taxid, name = "freq") |> 
    arrange(desc(freq))

  dat_dn_unfiltered <- dat_exp |>
    filter(`Abundance in Group 1` == "decreased") |> 
    count(last_taxname, last_taxid, name = "freq") |> 
    arrange(desc(freq))

  # Get taxa with freq >= min_freq in EITHER direction
  taxa_to_show <- unique(c(
    (dat_up_unfiltered |> filter(freq >= min_freq))$last_taxname,
    (dat_dn_unfiltered |> filter(freq >= min_freq))$last_taxname
  ))

  # Use provided p-values
  taxon_table_for_pval <- pval_dat

  # Get dominance and difference info
  taxa_stats <- bind_rows(
    dat_up_unfiltered |> mutate(direction = "increased"),
    dat_dn_unfiltered |> mutate(direction = "decreased")
  ) |>
    pivot_wider(
      id_cols = last_taxname,
      names_from = direction,
      values_from = freq,
      values_fill = 0
    ) |>
    mutate(
      difference = increased - decreased
    ) |>
    select(last_taxname, difference) |>
    left_join(taxon_table_for_pval, by = "last_taxname")

  # Combine all data
  result <- bind_rows(
    dat_up_unfiltered |> mutate(direction = "increased"),
    dat_dn_unfiltered |> mutate(direction = "decreased")
  ) |>
    filter(last_taxname %in% taxa_to_show) |>
    left_join(taxa_stats, by = "last_taxname") |>
    mutate(
      freq_display = freq,
      alpha_val = ifelse(pval > 0.05, 0.4, 1)
    ) |>
    # Create taxon factor ordered by number of signatures (descending)
    mutate(
      last_taxname = factor(last_taxname, 
        levels = names(sort(sapply(split(freq, last_taxname), sum), decreasing = TRUE)))
    )
  
  result
}

make_freq_plot <- function(dat_combined, title) {
  ggplot(dat_combined, aes(
    y = forcats::fct_rev(last_taxname), 
    x = freq_display,
    fill = direction,
    alpha = alpha_val
  )) +
    geom_col(position = "stack") +
    scale_x_continuous(labels = abs, position = "top") +
    scale_fill_manual(
      values = c("increased" = "#f1807e", "decreased" = "#6182a6"),
      labels = c("increased" = "Increased", "decreased" = "Decreased")
    ) +
    scale_alpha_identity() +
    labs(
      y = NULL, 
      x = NULL,
      title = title,
      fill = NULL
    ) +
    theme_minimal() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "top",
      plot.title = element_text(hjust = 0)
    )
}

make_taxon_table <- function(dat) {
  bugSigSimple::createTaxonTable(
    mutate(dat, comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs ")),
    n = Inf
  ) |> 
    select(
      `Taxon Name`,
      `Signatures` = total_signatures,
      `Increased` = increased_signatures,
      `Decreased` = decreased_signatures,
      `P-value` = `Binomial Test pval`
    ) |> 
    mutate(`Δ` = Increased - Decreased, .after = Decreased)
}

make_dt <- function(dat) {
  DT::datatable(
    dat,
    filter = 'top',
    class = 'row-border hover order-column nowrap',
    options = list(
      dom = 'lrtip',
      pageLength = 10
    )
  )
}

```

```{r load-data}

dat <- readRDS("example_data/perio_bugs.rds")

# Expand one row per bug and extract terminal taxon ----
dat_exp <- dat |> 
  unnest(c(`NCBI Taxonomy IDs`, `MetaPhlAn taxon names`)) |>
  mutate(
    last_taxname = get_most_specific(`MetaPhlAn taxon names`),
    last_taxid = get_most_specific(`NCBI Taxonomy IDs`)
  )

# Identify broad coverage signatures ----
broad_platforms <- c("DNA-DNA Hybridization", "Human Intestinal Tract Chip", 
                     "Illumina", "Roche454")

targeted_platforms <- c("RT-qPCR", "Non-quantitative PCR", "Sanger", "Mass spectrometry")

dat_broad <- dat |> 
  mutate(seq_keep = case_when(
    # use seq_type first if available
    `Sequencing type` %in% c("16S", "WMS") ~ TRUE,
    `Sequencing type` %in% c("PCR", "16S,PCR") ~ FALSE,
    # fall back to seq_plat if seq_type is NA
    is.na(`Sequencing type`) & `Sequencing platform` %in% broad_platforms ~ TRUE,
    is.na(`Sequencing type`) & `Sequencing platform` %in% targeted_platforms ~ FALSE,
    .default = NA
  )) |>
  filter(seq_keep)
```

# Background
Periodontitis is a chronic inflammatory disease affecting the supporting structures of teeth, including the gingiva, periodontal ligament, and alveolar bone. The disease is driven by dysbiosis of the oral microbiome, with shifts from commensal to pathogenic bacterial communities. Understanding which microbial taxa are consistently associated with periodontitis across multiple studies can help identify key drivers of disease and potential therapeutic targets.

PerioSigDB is a comprehensive database of periodontal microbial signatures curated from published literature, providing a resource for meta-analysis of microbiome studies. Here, we analyze periodontal disease-associated signatures from PerioSigDB to identify taxa that show consistent patterns of abundance changes across studies.

```{r signature-lengths}
#| echo: false
# Create dat_condition for signature-level analysis
dat_condition <- dat |> 
  mutate(comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs "))

dat_condition_broad <- dat_broad |> 
  mutate(comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs "))

# get signatures and lengths
allsigs <- bugsigdbr::getSignatures(dat_condition, tax.id.type = "taxname")
siglengths <- sapply(allsigs, length)

allsigs_broad <- bugsigdbr::getSignatures(dat_condition_broad, tax.id.type = "taxname")
siglengths_broad <- sapply(allsigs_broad, length)
```

# Broad Data Definition

Throughout this report, we present analyses for two data perspectives: "All Data" and "Broad Data Only." Broad data is defined as signatures from studies utilizing broad-coverage sequencing technologies, specifically:

  - **Sequencing type**: 16S ribosomal RNA or Whole Metagenome Sequencing (WMS)
  - **Sequencing platforms**: DNA-DNA Hybridization, Human Intestinal Tract Chip, Illumina, or Roche454

Signatures generated using targeted or limited approaches (RT-qPCR, non-quantitative PCR, Sanger sequencing, mass spectrometry) are excluded from the broad data subset. 

## Signature Length Distribution Comparison

```{r signature-length-comparison}
#| fig-width: 10
#| fig-cap: !expr "paste0('Distribution of signature lengths. Note: ', sum(siglengths > 50), ' signatures with length > 50 are excluded from this visualization (', sum(siglengths_broad > 50), ' from broad data).')"

# Combine data for visualization
sig_counts_all <- data.frame(siglength = siglengths) |>
  count(siglength, name = "count") |>
  mutate(data_type = "All Data")

sig_counts_broad <- data.frame(siglength = siglengths_broad) |>
  count(siglength, name = "count") |>
  mutate(data_type = "Broad Data Only")

# Combine data
sig_data <- bind_rows(sig_counts_all, sig_counts_broad)

# Create bar chart with overlaid bars
ggplot(sig_data, aes(x = siglength, y = count, fill = data_type)) +
  geom_col(position = "identity") +
  scale_fill_manual(values = c("All Data" = "#999999", "Broad Data Only" = "#2ecc71")) +
  scale_x_continuous(limits = c(0.5, 50.5), breaks = c(1, seq(5, 50, by = 5))) +
  labs(
    x = "Signature Length (number of taxa)",
    y = "Number of Signatures",
    title = "Distribution of Signature Lengths: All Data vs Broad Data Only",
    fill = "Data Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

This comparison illustrates how the broad data subset differs from the full dataset in terms of signature composition size. The overlaid bar chart shows the number of signatures at each length, with grey bars representing all signatures and green bars showing the broad data subset in front, allowing visual comparison of the distribution differences.

# Taxa Increased in Periodontitis

From the periodontal microbiome signatures, we generated a table of taxa showing consistent increases in periodontitis. The binomial test evaluates whether each taxon is more frequently reported as increased rather than decreased across studies. Taxa appearing in this table show statistically significant directional consistency.

```{r create-taxon-tables}
#| cache: true
#| echo: false 
taxon_table <- make_taxon_table(dat)
taxon_table_broad <- make_taxon_table(dat_broad)
```

```{r extract-pvals}
#| echo: false
# Extract p-values for use in frequency plots
pval_dat_all <- taxon_table |>
  select(last_taxname = `Taxon Name`, pval = `P-value`)

pval_dat_broad <- taxon_table_broad |>
  select(last_taxname = `Taxon Name`, pval = `P-value`)
```

## Frequency Distribution

Bars show the count of signatures in which each taxon appears as increased or decreased. Excluded taxa appearing in fewer than 3 signatures. Transparency indicates statistical significance (opaque: p ≤ 0.05; transparent: p > 0.05).

::: {.panel-tabset}

### All Data

:::: {style="max-height: 800px; overflow-y: auto; border: 1px solid #ddd; padding: 10px;"}

```{r freq-increased-all}
#| fig-width: 10
#| fig-height: 15

dat_freq_combined <- make_freq_dat_combined(dat_exp, pval_dat_all) 

dat_freq_combined |>
  filter(difference > 0) |>
  arrange(pval) |>
  slice_head(n=40) |>
  mutate(direction = factor(direction, levels = c("increased", "decreased"))) |>
  make_freq_plot("Taxon Frequency: Increased > Decreased")
```

::::

### Broad Data Only

:::: {style="max-height: 800px; overflow-y: auto; border: 1px solid #ddd; padding: 10px;"}

```{r freq-increased-broad}
#| fig-width: 10
#| fig-height: 12

dat_exp_broad <- dat_broad |> 
  unnest(c(`NCBI Taxonomy IDs`, `MetaPhlAn taxon names`)) |>
  mutate(
    last_taxname = get_most_specific(`MetaPhlAn taxon names`),
    last_taxid = get_most_specific(`NCBI Taxonomy IDs`)
  )

dat_freq_broad_combined <- make_freq_dat_combined(dat_exp_broad, pval_dat_broad) 
dat_freq_broad_combined |>
  filter(difference > 0) |>
  mutate(direction = factor(direction, levels = c("increased", "decreased"))) |>
  make_freq_plot("Taxon Frequency: Increased > Decreased")
```

::::

:::

## Taxon Summary Tables

::: {.panel-tabset}

### All Data

```{r taxon-summary-increased-all}
taxon_table |> 
  filter(Increased > Decreased) |>
  arrange(`P-value`) |>
  make_dt()
```

### Broad Data Only

```{r taxon-summary-increased-broad}
taxon_table_broad |> 
  filter(Increased > Decreased) |>
  arrange(`P-value`) |>
  make_dt()
```

:::

# Taxa Decreased in Periodontitis

Conversely, we identified taxa that are consistently depleted in periodontitis compared to healthy controls. These may represent commensal or health-associated microbes that are displaced during disease progression.

## Frequency Distribution

Bars show the count of signatures in which each taxon appears as increased or decreased. Excluded taxa appearing in fewer than 3 signatures. Transparency indicates statistical significance (opaque: p ≤ 0.05; transparent: p > 0.05).

::: {.panel-tabset}

### All Data

:::: {style="max-height: 800px; overflow-y: auto; border: 1px solid #ddd; padding: 10px;"}

```{r freq-decreased-all}
#| fig-width: 10
#| fig-height: 15

dat_freq_combined |> 
  filter(difference < 0) |>
  arrange(pval) |>
  slice_head(n=40) |>
  mutate(freq_display = abs(freq_display)) |>
  mutate(direction = factor(direction, levels = c("decreased", "increased"))) |>
  make_freq_plot("Taxon Frequency: Decreased > Increased")
```

::::

### Broad Data Only

:::: {style="max-height: 800px; overflow-y: auto; border: 1px solid #ddd; padding: 10px;"}

```{r freq-decreased-broad}
#| fig-width: 10
#| fig-height: 12

dat_freq_broad_combined |>
  filter(difference < 0) |>
  mutate(freq_display = abs(freq_display)) |>
  mutate(direction = factor(direction, levels = c("decreased", "increased"))) |>
  make_freq_plot("Taxon Frequency: Decreased > Increased")
```

::::

:::

## Taxon Summary Tables

::: {.panel-tabset}

### All Data

```{r taxon-summary-decreased-all}

taxon_table |> 
  filter(Decreased >= Increased) |>
  arrange(`P-value`) |>
  make_dt()
```

### Broad Data Only

```{r taxon-summary-decreased-broad}

taxon_table_broad |>
  filter(Decreased >= Increased) |>
  arrange(`P-value`) |>
  make_dt()
```

:::

# Cluster Analysis

To assess consistency across studies, we calculated pairwise Jaccard similarities between all signatures. The heatmap below shows how similar signatures are to each other, with hierarchical clustering grouping similar signatures together. The top annotation bar shows the size (number of taxa) of each signature.

```{r cluster-analysis}
#| fig-width: 10
#| fig-height: 8


# require length > 1
allsigs <- allsigs[sapply(allsigs, length) > 1]
siglengths <- sapply(allsigs, length)

mydists <- BugSigDBStats::calcPairwiseOverlaps(allsigs)

# Create a matrix of Jaccard similarities (0 for no overlap, 1 for 100% overlap)
jmat <- BugSigDBStats::calcJaccardSimilarity(allsigs)

# Remove annotation name, use column title on heatmap
ha <- HeatmapAnnotation(
  `Signature Length` = anno_barplot(siglengths),
  show_annotation_name = FALSE  # Hide the annotation label
)
hm <- Heatmap(
  jmat,
  top_annotation = ha, 
  column_title = "Signature Length",  # Add as column title
  column_title_side = "top",
  left_annotation = NULL,
  row_names_max_width = unit(20, "cm"),
  column_names_max_height = unit(20, "cm"),
  row_labels = gsub('^.+?_(UP|DOWN)$', "\\1", rownames(jmat)),
  column_labels = gsub('^.+?_(UP|DOWN)$', "\\1", colnames(jmat)),
  row_names_gp = gpar(fontsize = 6),  
  column_names_gp = gpar(fontsize = 6),
  show_column_dend = FALSE,
  heatmap_legend_param = list(
    title = 'Jaccard Similarity',
    direction = "horizontal"  # Make legend horizontal
  )
)

# Draw with legend at bottom
draw(hm, heatmap_legend_side = "bottom")
```

 Signatures reporting increased taxa (UP) in periodontitis show particularly strong clustering, suggesting a core set of disease-associated microbes is consistently identified.
 
# Explore Species Within Clusters
```{r create-taxa-heatmap}
#| fig-width: 10
#| fig-height: 8

# Helper function to get author labels (without year)
get_author_label <- function(dat, sname) {
  id <- unlist(strsplit(sname, "_"))[1]
  id <- gsub("^bsdb:", "", id)
  id <- unlist(strsplit(id, "/"))
  
  sdat <- dat[dat$Study == paste0("Study ", id[1]) & 
                dat$Experiment == paste0("Experiment ", id[2]), ]
  
  if(nrow(sdat) == 0) return("Unknown")
  
  author <- sdat$`Authors list`[1]
  # Get first author last name
  aspl <- unlist(strsplit(author, " "))[1]
  year <- sdat$Year[1]
  
  # Add "et al." if multiple authors
  if(grepl(",", author)) {
    return(paste0(aspl, " et al. (", year, ")"))
  } else {
    return(paste0(aspl, " (", year, ")"))
  }
}
# Helper function to plot frequency heatmaps
create_taxa_heatmap <- function(dat_condition, 
                                clust_up, 
                                jmat, 
                                hc, 
                                tax_level = "species") {
  
  # Get signatures
  sigs_taxa <- bugsigdbr::getSignatures(
    dat_condition,
    tax.id.type = "taxname",
    tax.level = tax_level,
    exact.tax.level = FALSE
  )
  
  # Filter to only signatures in clust_up
  sigs_taxa <- sigs_taxa[names(sigs_taxa) %in% clust_up]
  
  # Remove empty signatures
  sigs_taxa <- sigs_taxa[sapply(sigs_taxa, length) > 0]
  
  # Create binary presence/absence matrix
  all_taxa <- unique(unlist(sigs_taxa))
  
  taxa_matrix <- matrix(0, 
                        nrow = length(sigs_taxa),
                        ncol = length(all_taxa),
                        dimnames = list(names(sigs_taxa), all_taxa))
  
  for(i in seq_along(sigs_taxa)) {
    sig_name <- names(sigs_taxa)[i]
    taxa_in_sig <- sigs_taxa[[i]]
    taxa_matrix[sig_name, taxa_in_sig] <- 1
  }
  
  # Calculate taxa frequency
  taxa_freq <- colSums(taxa_matrix) / nrow(taxa_matrix)
  taxa_order <- order(taxa_freq, decreasing = TRUE)
  taxa_freq_ordered <- taxa_freq[taxa_order]
  
  # Re-cluster using original distance matrix
  sigs_to_use <- rownames(taxa_matrix)
  jmat_subset <- jmat[sigs_to_use, sigs_to_use]
  hc_subset <- hclust(dist(jmat_subset), method = hc$method)
  dend_subset <- as.dendrogram(hc_subset)
  
  # Reorder taxa_matrix to match clustering
  taxa_matrix_ordered <- taxa_matrix[hc_subset$labels, taxa_order]
  
  # Apply author labels
  sig_labels_ordered <- vapply(rownames(taxa_matrix_ordered), 
                               get_author_label, 
                               character(1), 
                               dat = dat_condition)
  
  labels(dend_subset) <- sig_labels_ordered
  rownames(taxa_matrix_ordered) <- sig_labels_ordered
  
  # Create color matrix
  color_matrix <- taxa_matrix_ordered
  for(j in 1:ncol(color_matrix)) {
    for(i in 1:nrow(color_matrix)) {
      if(color_matrix[i, j] == 0) {
        color_matrix[i, j] <- NA
      } else {
        color_matrix[i, j] <- taxa_freq_ordered[j]
      }
    }
  }
  
  # Color function
  col_fun <- colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
  
  # Create heatmap
  hm_taxa <- Heatmap(
    color_matrix,
    name = "Frequency",
    col = col_fun,
    na_col = "white",
    cluster_rows = dend_subset,
    cluster_columns = FALSE,
    show_row_dend = TRUE,
    row_dend_width = unit(1, "cm"),
    row_names_side = "left",
    column_names_side = "bottom",
    row_names_gp = gpar(fontsize = 10),
    column_names_gp = gpar(fontsize = 10),
    column_names_rot = 90,
    column_title = NULL,
    row_title = NULL,
    top_annotation = NULL,
    heatmap_legend_param = list(
      title = "Frequency",
      direction = "vertical",
      title_gp = gpar(fontsize = 9),
      labels_gp = gpar(fontsize = 8),
      at = c(0, 0.25, 0.5, 0.75, 1),
      labels = c("0", "0.25", "0.5", "0.75", "1"),
      legend_height = unit(2, "cm"),
      legend_width = unit(0.3, "cm")
    )
  )
  
  return(hm_taxa)
}
```

Another way to explore taxa associated with periodontitis is to examine which genera/species are present across the cluster of highly similar signatures identified above. The heatmap shows each signature (rows) and taxa (columns, ordered by frequency), with colors indicating the overall frequency of each taxon across all signatures in the cluster.


```{r heatmap-genus}
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Genus-level signatures: The heatmap above shows genus-level taxa present across signatures in the high-similarity cluster."

# identify the clusters
hc <- hclust(dist(jmat))

#set the number of clusters here with k
clusts <- sort(cutree(hc, k = 2)) 
clusts <- lapply(unique(clusts), function(i) names(clusts)[clusts == i])

# proceed with cluster of UP signatures with high similarity
clust_up <- clusts[[2]]

hm_genus <- create_taxa_heatmap(dat_condition, clust_up, jmat, hc, 
                                  tax_level = "genus")
draw(hm_genus, heatmap_legend_side = "right")
```


```{r heatmap-species}
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Species-level signatures: For a more detailed view, the heatmap above shows the same signatures at the species level."

hm_species <- create_taxa_heatmap(dat_condition, clust_up, jmat, hc, 
                                  tax_level = "species")
draw(hm_species, heatmap_legend_side = "right")

```
# References

:::{#refs}
:::